---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
set.seed(42)

population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

population
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sampled_data <- population |> 
  slice_sample(n=20)

sampled_data 
```


### Task 4: Calculate the mean age and standard deviation of the sample
```{r}
sampled_stats <- sampled_data |> 
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )

sampled_stats
```


### Task 5: Calculate the mean age and standard deviation of the population
```{r}
population_stats <- population |> 
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )

population_stats

```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

ANSWER: They are similar, but the population stats has a slightly higher median age and a slightly smaller standard deviation. The sample median age is 49.02, but the population age is 49.873. And the standard deviation for the sampled dataset is 18.26847, while it is 17.50627 for the population. When using a smaller sample size of 20, the median age happened to get closer to the population age, its standard deviation increased to 19.6251. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
